Depth and thickness of tectonic tremor in the northeastern Olympic Peninsula

### 2021-06-11 paper/agujournaltemplate.tex
New modifs on paper

\begin{keypoints}
\item We use seismic data from small-aperture arrays in the Olympic Peninsula, Washington.
\item We compute the depth of the tremor source from the stacking of horizontal-to-vertical cross-correlations of seismic data.
\item Tremor is located close to the plate boundary in a region no more than 2-3 kilometers thick.
\end{keypoints}

@@

The occurrence of tremor seems to be linked to low effective normal stress or high fluid pressure near the location of the source of the tremor. Indeed, ~\citeA{SHE_2006} have observed a high ratio between P-wave velocity and S-wave velocity in the subducting oceanic crust near the location of the LFEs in western Shikoku, Japan. In Cascadia, ~\citeA{AUD_2009} have computed receiver functions of teleseismic waves in Vancouver Island, and analyzed the delay times between the forward-scattered P-to-S, and back-scattered P-to-S and S-to-S conversions at two seismic reflectors identified as the top and bottom of the oceanic crust. It allowed them to compute the P-to-S velocity ratio of the oceanic crust layer and the S-wave velocity contrast at both interfaces. The very low Poisson's ratio of the layer could not be explained by the mineral composition, and they interpreted it as evidence for high pore-fluid pressure. The link between tremor and high pore fluid pressure is also supported by the influence of tidal cycles on the variations of tremor occurrence. \citeA{NAK_2008} noticed that tremor swarms often exhibit occurrences with a periodicity of about 12 or 24 h, and concluded that they are probably related to Earth tides. Their occurrence is also well correlated with time evolution of Coulomb failure stress (CFS) and CFS rate. However, they noted that tremor occurrences are advanced by a few hours relative to CFS, from which they conclude that a simple Coulomb threshold model is not sufficient to explain tremor occurrence. Instead they point out that the correlation of tremor occurrence and the CFS rate as well as the time delay between both could be reproduced by using the rate- and state-dependent friction law. ~\citeA{THO_2009} have also observed that tremor occurrence on the deep San Andreas fault are correlated with small, tidal shear stress changes. They explain it by a very weak fault zone with low effective normal stress, probably due to near-lithostatic pore pressures at the depth of the tremor source region. \\

@@

Moreover, the zone with high low Poisson's ratio observed by ~\citeA{HYN_2015} has a large vertical extent (about 10 kilometers). If the whole zone is associated with quartz deposition and tectonic tremor generation, we should also observe a large vertical distribution of the source of the tremor. ~\citeA{KAO_2006} have used a Source Scanning Algorithm to detect and locate tremor, and have indeed located tremor in the continental crust, with a wide depth range of over 40 km. They noted that this wide depth range could not arise from either analysis uncertainties or a systematic bias in the velocity model they used. Uncertainties on the location of the tremor have been estimated by ~\citeA{IDE_2012} at about 1.5 km in the horizontal direction and 4.5 km in the vertical direction. A follow-up study by ~\citeA{KAO_2009} gave a thickness of the tremor zone of 5-10 km. This depth range is inconsistent with the depth of the LFEs, which have been located on a thin band at or near the plate interface with a rupture mechanism that corresponds to the thrust dip angle ~\cite{IDE_2007_GRL}. In the Olympic Peninsula, LFE families have been identified and located by ~\citeA{CHE_2017_JGR}, and all LFE families were located near the plate interface. Further study is thus needed to narrow the uncertainty on the depth of the source of the tremor, and verify whether tremor could occur in a wider zone than LFEs. \\

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth, trim={0cm 2.5cm 0cm 9.5cm},clip]{figures/arrays_location.eps}
\caption{Map showing the location of the eight arrays (black triangles) used in this study and tremor locations (grey dots) located using these arrays ~\cite{GHO_2012}. Inset shows the study area with the box marking the area covered in the main map. Contour lines represent a model of the depth of the plate interface ~\cite{MCC_2006}.}
\label{pngfiguresample}
\end{figure}

@@

For each array, and every 5 km by 5 km grid cell located within 25 km, we analyze the one-minute-long time windows corresponding to all the tremor epicenters located within the grid cell. For each one-minute-long time window, we downloaded the seismic data for each 3-component seismic station of the array. Then, for each seismic station and each channel, we detrended the data, tapered the first and last 5 seconds of the data with a Hann window, removed the instrument response, bandpass filtered between 2 and 8 Hz, and resampled the data to 20 Hz. All these preprocessing operations were done with the Python package obspy. For each seismic station and each one-minute-long time window, we cross correlated the vertical component with the East-West component and with the North-South component. Then, we stacked the cross correlation functions over all the seismic stations of the array. At each step of this method we experimented with a linear stack, a nth-root stack, and a phase-weighted stack (~\cite{SCH_1997} and found that the phase-weighted stack worked best so all the stacks discussed here used the phase-weighted method. We then computed the envelopes of the stacked cross-correlation functions. Figure 2 shows an example of the envelopes of the cross-correlation functions for the Big Skidder array for the 110 one-minute-long time windows when tremor was detected in a 5 km by 5 km grid cell located 7 kilometers southwest of the array. We can see that for about half of the tremor windows, there is a peak in the cross-correlation at about 4.85 s. As the energy of the P-waves is expected to be higher on the vertical component, and the energy of the S-waves to be higher on the horizontal components, we assume that this peak corresponds to the time lag between the arrival of a direct P-wave and a direct S-wave. We then stacked the cross-correlation functions over all the one-minute-long time windows. We assume that the time of the maximum absolute value of the peak of the stack is the time lag between the arrival of the direct P-wave and the arrival of the direct S-wave.

\begin{figure}
\noindent\includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm},clip]{figures/BS_-05_-05_PWS_PWS_cluster_ccwin.eps}
\caption{Envelopes of the stacked cross-correlation functions ordered by time for the Big Skidder array for the 110 one-minute-long time windows when tremor was detected in a 5 km by 5 km grid cell located 7 kilometers southwest of the array. Top time windows in black are the time windows that fit well with the stack. Bottom time windows in blue are time windows that do not fit well with the stack. Left panel is the cross-correlation of the EW component with the vertical component, and right panel is the cross-correlation of the NS component with the vertical component. The cross-correlation functions have been cut between 2 and 8 seconds to focus on the time lags around the peaks}
\label{pngfiguresample}
\end{figure}

Only about half of the envelopes of the cross-correlation functions have a distinct peak that coincides with the peak in the stacked cross-correlation. The other cross-correlations functions show either a distinct peak at another time lag, or no clearly visible peak. This may be either because the source of the tremor during the corresponding one-minute-long time window was mislocated, or because the signal-to-noise ratio is too low. To improve the signal-to-noise ratio of the peak in the stacked cross-correlation, we divided the one-minute-long time windows into two clusters, the ones that match well the stacked cross-correlations, and the ones that do not match it well. For each one-minute-long time window, we took the envelope of the stack of cross-correlation functions between the horizontal and the vertical components of each station in an array.  We then cross-correlated it with the stack over all the envelopes. From this last cross-correlation function, we then computed its value at zero lag time, its maximum absolute value, and the time lag at which it takes its maximum absolute value and the ratio of the maximum absolute value to the rms of the correlation function.  We considered lag times from X to Y in each of these measures. We did this for both the East-West component and the North-South component. Each one-minute-long time window is thus associated with eight values of quality criteria. We then classified each one-minute-long time window into two different clusters, based on the value of these criteria, using a K-means clustering algorithm (function sklearn.cluster.KMeans from the Python library SciKitLearn). The K-means procedure is as follows: We choose the number of clusters R, then we arbitrarily choose a center for each cluster. We put each one-minute-long time window into the cluster to which it is closest (based on the values of the eight criteria). Once all one-minute-long time windows have been put in a cluster, we recompute the mean of the eight criteria for each cluster, and iterate the procedure until convergence. On average, about 35\% of the time windows fit well with the stack and are kept in the first cluster, while 65\% do not fit well with the stack and are removed. For each cluster, we then stacked the cross-correlation functions over all the one-minute-long time windows belonging to the cluster using a phase-weighted stack. We always use two clusters. We try using more clusters, but it did not improve the final stack. Figure 3 shows the stack of the envelopes, as described above, for each of the two clusters as well as a stack over all windows of the stack of cross correlation functions. The clustering has greatly improved the amplitude of the peak for one of the clusters, and made the peak nearly disappear for the other cluster.

\begin{figure}
\noindent\includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm},clip]{figures/BS_-05_-05_PWS_PWS_cluster_stackcc.eps}
\caption{Stack of the cross-correlation functions over all the 110 time windows from Figure 2. We used a phase-weighted stack for the stack over the stations and for the stack over the time windows. The black line is the stack over all time windows of the stack of cross-correlation functions over all stations within the array. The red line on the left panels is the stack over all the time windows in the first cluster of the envelopes of the stack of cross-correlation functions over stations within the array.  This corresponds to time windows that do not fit well with the stack. The blue lines on the right panels are the same as the red lines, except it uses time windows in the second cluster (which contains the time windows that fit well with the stack). Top panels are the cross-correlation of the EW component with the vertical component, and bottom panels are the cross-correlation of the NS component with the vertical component. The stacked cross-correlation function has been cut between 2 and 8 seconds to focus on the time lags around the peak.}
\label{pngfiguresample}
\end{figure}

We did this analysis for grid cells located in a 50 km by 50 km area centered on each of the eight arrays. We thus consider 11 * 11 * 8 = 968 values of the time lag between the arrival of the direct P-wave and the arrival of the direct S-wave. We first assumed that the epicenter is at the center of the grid and determined the depth that satisfies that epicenter and the observed S-minus-P. The velocity models are taken from the 3D velocity model from ~\citeA{MER_2020}. For each array, we looked for the closest grid point (in terms of latitude and longitude) in the 3D model by ~\citeA{MER_2020} and used the corresponding layered model of compressional and shear wave speeds to compute the depth. We thus used 8 different 1D velocity models corresponding to the 8 arrays. This allows us to take into account the substantial variations of the Poisson's ratio in the East-West direction.

>>>>

\begin{keypoints}
\item We use seismic data from small-aperture arrays in the Olympic Peninsula, Washington and tremor epicenters determined previously by a multibeam backprojection method. 
\item We compute the depth of the tremor source from estimates of S minus P times determined from stacking horizontal-to-vertical cross-correlations of seismic data.
\item Tremor is located close to the plate boundary in a region no more than 2-3 kilometers thick.
\end{keypoints}

@@

The occurrence of tremor seems to be linked to low effective normal stress or high fluid pressure near the location of the tremor. Indeed, ~\citeA{SHE_2006} have observed a high P-wave to S-wave velocity ratio in the subducting oceanic crust near the location of the LFEs in western Shikoku, Japan. In Cascadia, ~\citeA{AUD_2009} have computed teleseismic receiver functions in Vancouver Island, and analyzed the delay times between the forward-scattered P-to-S, and back-scattered P-to-S and S-to-S conversions at two seismic reflectors identified as the top and bottom of the oceanic crust. It allowed them to compute the P-to-S velocity ratio of the oceanic crust layer and the S-wave velocity contrast at both interfaces. The very low Poisson's ratio of the layer could not be explained by the mineral composition, and they interpreted it as evidence for high pore-fluid pressure. The link between tremor and high pore fluid pressure is also supported by the influence of tidal cycles on the variations of tremor occurrence. \citeA{NAK_2008} noticed that tremor swarms often exhibit occurrences with a periodicity of about 12 or 24 h, and concluded that they are probably related to Earth tides. Their occurrence is also well correlated with time evolution of Coulomb failure stress (CFS) and CFS rate. However, they noted that tremor occurrences are advanced by a few hours relative to CFS, from which they conclude that a simple Coulomb threshold model is not sufficient to explain tremor occurrence. Instead they point out that the correlation of tremor occurrence and the CFS rate as well as the time delay between both could be reproduced by using the rate- and state-dependent friction law. ~\citeA{THO_2009} have also observed that tremor occurrence on the deep San Andreas fault are correlated with small, tidal shear stress changes. They explain it by a very weak fault zone with low effective normal stress, probably due to near-lithostatic pore pressures at the depth of the tremor source region. \\

@@

Moreover, the zone with high low Poisson's ratio observed by ~\citeA{HYN_2015} has a large vertical extent (about 10 kilometers). If the whole zone is associated with quartz deposition and tectonic tremor generation, we should also observe a large vertical distribution of the source of the tremor. ~\citeA{KAO_2006} have used a Source Scanning Algorithm to detect and locate tremor, and have indeed located tremor in the continental crust, with a wide depth range of over 40 km. They noted that this wide depth range could not arise from either analysis uncertainties or a systematic bias in the velocity model they used. Uncertainties on the location of the tremor have been estimated by  ~\citeA{IDE_2012} at about 1.5 km in epicenter and 4.5 km in depth. A follow-up study by ~\citeA{KAO_2009} gave a thickness of the tremor zone of 5-10 km. This depth range is inconsistent with the depth of the LFEs, which have been located on a thin band at or near the plate interface with a rupture mechanism that corresponds to the thrust dip angle ~\cite{IDE_2007_GRL}. In the Olympic Peninsula, LFE families have been identified and located by ~\citeA{CHE_2017_JGR}, and all LFE families were located near the plate interface. Further study is thus needed to narrow the uncertainty on the depth of the source of the tremor, and verify whether tremor could occur in a wider zone than LFEs. \\

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth, trim={0cm 2.5cm 0cm 9.5cm},clip]{figures/arrays_location.eps}
\caption{Map showing the location of the eight arrays (black triangles) used in this study and tremor locations (grey dots) located using these arrays ~\cite{GHO_2012}. Inset shows the study area with the box marking the area covered in the main map. Contours represent a model of the depth of the plate interface ~\cite{MCC_2006}.}
\label{pngfiguresample}
\end{figure}

@@

For each array, and every 5 km by 5 km grid cell located within 25 km, we analyze the 1-minute long time windows corresponding to all the tremor epicenters located within the grid cell. For each three-component seismic station and each channel, we detrended the data, tapered the first and last 5 seconds with a Hann window, removed the instrument response, bandpass filtered between 2 and 8 Hz, and resampled the data to 20 Hz. All these preprocessing operations were done with the Python package obspy. For each seismic station and each one-minute-long time window, we cross correlated the vertical component with the East-West component and with the North-South component. Then, we stacked the cross correlation functions over all the seismic stations of the array. We call this stack $S_{i j} (\tau)$ for time lag $\tau$, the $i$-th channel ($i$=1 or 2 for the North or East channel) and the $j$-th one-minute time window for a given epicenter grid/array pair. At each step of this method we experimented with a linear stack, a $n$th-root stack, and a phase-weighted stack ~\cite{SCH_1997} and found that the phase-weighted stack worked best. So, all the stacks discussed here used the phase-weighted method.  Figure 2 shows an example of the envelopes of $S_{i j} (\tau)$ for the Big Skidder array for the 110 one-minute long time windows when tremor was detected in the 5 km by 5 km grid cell located 7 kilometers southwest of the array. As the energy of the P-waves is expected to be higher on the vertical component, and the energy of the S-waves to be higher on the horizontal components, we interpret this peak to correspond to the time lag between the arrival of the direct P- and S-waves.

\begin{figure}
\noindent\includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm},clip]{figures/BS_-05_-05_PWS_PWS_cluster_ccwin.eps}
\caption{Envelopes of the stacked cross-correlation functions ordered by time for the Big Skidder array for the 110 one-minute long time windows when tremor was detected in a 5 km by 5 km grid cell located 7 kilometers southwest of the array. Top shows time windows (black) that fit the stack well. Bottom shows time windows (blue) that do not fit the stack well. Left panel is the cross-correlation of the EW component with the vertical component, and right panel is the cross-correlation of the NS component with the vertical component. The cross-correlation functions have been cut between 2 and 8 seconds to focus on the time lags around the peaks.}
\label{pngfiguresample}
\end{figure}

Only about half of the envelopes of the cross-correlation functions have a distinct peak that coincides with the peak in the stacked cross-correlation. The other cross-correlation functions show either a distinct peak at another time lag, or no clearly visible peak. This may be either because the tremor epicenter was mislocated, or because the signal-to-noise ratio is too low. \\

To improve the signal-to-noise ratio of the peak in the stacked cross-correlation, we divided the one-minute-long time windows into two clusters, the ones that match the stacked cross-correlations well, and the ones that do not. In order to do the clustering, we stacked the stacked cross-correlation functions, $S_{i j} (\tau)$, over all the one-minute-long time windows to obtain $S_i (\tau)$. For each 1-minute time window $j$ we cross-correlated $S_i (\tau)$ with $S_{i j} (\tau)$ to obtain $S^c_{i j} (\tau)$.  For each $i$ and $j$ we determine 4 numbers from $S^c_{i j} (\tau)$: its value at zero lag time, its maximum absolute value, the time lag at which it takes its maximum absolute value and the ratio of the maximum absolute value to the rms of the correlation function. We considered lag times from X to Y in each of these measures. Each one-minute-long time window is thus associated with eight values of quality criteria, four for each of the the East and North channels. We then classified each one-minute-long time window into two different clusters, based on the value of these criteria, using a K-means clustering algorithm (function sklearn.cluster.KMeans from the Python library SciKitLearn). The K-means procedure is as follows: We choose to have 2 clusters, then we arbitrarily choose a center for each cluster. We put each one-minute-long time window into the cluster to which it is closest (based on the values of the eight criteria). Once all one-minute-long time windows have been put in a cluster, we recompute the mean of the eight criteria for each cluster, and iterate the procedure until convergence. On average, about 35\% of the time windows fit well with the stack and are kept in the first cluster, while 65\% do not fit well with the stack and are removed. For each cluster, we then stacked the cross-correlation functions over all the one-minute-long time windows belonging to that cluster using a phase-weighted stack. We tried using more than 2 clusters, but it did not improve the final stack of the best cluster. Figure 3 shows the envelope of the stack of $S_{i j} (\tau)$ over the time windows $j$ that correspond to cluster 0 (red) and cluster 1 (blue).  The clustering has greatly improved the amplitude of the peak for cluster 1, and made the peak nearly disappear for cluster 0.

\begin{figure}
\noindent\includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm},clip]{figures/BS_-05_-05_PWS_PWS_cluster_stackcc.eps}
\caption{Stack of the cross-correlation functions over all the 110 time windows from Figure 2. The black line is the stack over all time windows of the stack of cross-correlation functions over all stations within the array. The red line on the left panels is the stack over all the time windows in the first cluster of the envelopes of the stack of cross-correlation functions over stations within the array.  This corresponds to time windows that do not fit well with the stack. The blue lines on the right panels are the same as the red lines, except it uses time windows in the second cluster (which contains the time windows that fit well with the stack). Top panels are the cross-correlation of the EW component with the vertical component, and bottom panels are the cross-correlation of the NS component with the vertical component. The stacked cross-correlation function has been cut between 2 and 8 seconds to focus on the time lags around the peak.}
\label{pngfiguresample}
\end{figure}

We did this analysis for grid cells located in a 50 km by 50 km area centered on each of the eight arrays. We thus consider up to 11 * 11 * 8 = 968 values of the time lag between the direct P-wave and S-wave. We assumed that the epicenter is at the center of the grid and determined the depth that satisfies that epicenter and the observed S-minus-P time. The velocity models are taken from the 3D velocity model from ~\citeA{MER_2020}. For each array, we looked for the closest grid point (in terms of latitude and longitude) in the 3D model by ~\citeA{MER_2020} and used the corresponding layered model of compressional and shear wave speeds to compute the tremor depth. We thus used 8 different 1D velocity models corresponding to the 8 arrays. This allows us to take into account the substantial variations of the Poisson's ratio in the East-West direction.
